---
title: "Long Acting Octreotide for Prevention of Variceal Bleeding"
author: "Sammie Rosen and Keith Hazleton"
date: "`r Sys.Date()`"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Variceal hemorrhage poses a significant threat to children with chronic liver disease, often leading to severe morbidity and mortality. Octreotide, a somatostatin-analogue, reduces portal venous flow, decreasing blood flow to varices associated with portal hypertension. This retrospective cohort study at Children’s Hospital Los Angeles (CHLA) identified 44 patients who received at least one dose of subcutaneous long-acting octreotide between January 2013 and December 2023. Data on variceal grades, EGD interventions, bleeding episodes, and adverse events were collected.

# Data Loading and Setup

```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(dplyr)
library(ggsignif)
library(lubridate)
library(lmerTest)  # For linear mixed models
library(ggpubr)  
library(lme4)
library(gridExtra)
library(gtsummary)

# Set working directory

setwd("~/Library/CloudStorage/OneDrive-ChildrensHospitalLosAngeles/Octreotide Study/")

# Load Datasets

datatable <- read.csv("data/csv files/cleaned_datatable.csv")
labsdata <- read.csv("data/csv files/labsdata.csv")
endoscopydata <- read.csv("data/csv files/Endoscopy Reports(updated).csv")
dosedata <- read.csv("data/csv files/oct_doses_withrand.csv")


#### setting up the data #### 
# Convert the date columns to Date type
datatable$date.of.first.dose <- as.Date(datatable$date.of.first.dose, format="%Y-%m-%d")
endoscopydata$Date.of.endoscopy <- as.Date(endoscopydata$Date, format="%m/%d/%Y")

# Get unique rand values from data
unique_rands <- unique(datatable$rand)

# Merge the datasets
merged_data <- endoscopydata %>%
  left_join(datatable, by = c("Rand_id" = "rand"))

# Ensure Grade.of.varices is numeric, coercing any non-numeric values to NA
merged_data$Grade.of.varices <- as.numeric(merged_data$Grade.of.varices)

# Filter the dates
before_filtered_data <- merged_data %>%
  filter(Date.of.endoscopy < date.of.first.dose)
 
after_filtered_data <- merged_data %>%
  filter(Date.of.endoscopy >= date.of.first.dose)

no_scopes <- datatable%>%
  filter(Number.of.Endoscopies == 0)

####Process the Data for Before and After Values ####

# 1. Find the "Before" value closest to the treatment date for each patient
before_data <- merged_data %>%
  filter(Date.of.endoscopy < date.of.first.dose) %>%
  group_by(Rand_id) %>%
  filter(Date.of.endoscopy == max(Date.of.endoscopy)) %>%  # Get the closest date before treatment
  select(Rand_id, Grade.of.varices, Date.of.endoscopy, age.at.first.dose, Disease.State)

# 2. Calculate the mean "After" grade for each patient
after_data <- merged_data %>%
  filter(Date.of.endoscopy >= date.of.first.dose) %>%
  group_by(Rand_id) %>%
  summarise(After_Grade = mean(Grade.of.varices, na.rm = TRUE))  # Average after treatment

# 3. Merge before and after data by Rand_id
combined_data <- left_join(before_data, after_data, by = "Rand_id") %>%
  filter(!is.na(After_Grade))  # Ensure both before and after values are present
#### Create a long format for boxplot ####
processed_data_long <- combined_data %>%
  pivot_longer(
    cols = c("Grade.of.varices", "After_Grade"),
    names_to = "TimePeriod",
    values_to = "Grade_of_Varices"
  ) %>%
  mutate(TimePeriod = factor(TimePeriod, levels = c("Grade.of.varices", "After_Grade"),
                             labels = c("Before", "After")))

# Add the 'AgeGroup' column to processed_data_long
processed_data_long$AgeGroup <- cut(processed_data_long$age.at.first.dose,
                                    breaks = seq(0, 25, by = 5), # Creating 5-year intervals up to 25
                                    labels = c("0-4", "5-9", "10-14", "15-19", "20-25"),
                                    right = FALSE)  # right = FALSE means the interval is [0, 5), [5, 10), etc.

```

# Data Visualization

```{r}
#### Boxplot with statistical test and colored spaghetti lines by AgeGroup ####
combined_boxplot_with_stats <- ggplot(processed_data_long, aes(x = TimePeriod, y = Grade_of_Varices, fill = TimePeriod)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA, width = 0.6) +  # Boxplot with fill aesthetic
  geom_line(aes(group = Rand_id, color = AgeGroup), size = 1.2, alpha = 0.8) +  # Spaghetti lines colored by AgeGroup
  geom_jitter(aes(color = TimePeriod), width = 0, size = 2, shape = 21, alpha = 0.7) +  # Jittered points by TimePeriod
  scale_fill_manual(values = c("Before" = "gray80", "After" = "gray50")) +  # Boxplot fill colors
  scale_color_manual(values = c("Before" = "black", "After" = "darkgray")) +  # Point colors
  scale_color_manual(values = c("0-4" = "gray90", "5-9" = "gray70", "10-14" = "gray50", 
                                "15-19" = "gray30", "20-25" = "gray10")) +  # Age group line colors
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),            # Remove grid lines
    axis.line = element_line(color = "black"),# Add axis lines
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "right",                # Position the legend on the right
    legend.title = element_text(size = 12),   # Adjust the legend title size
    legend.text = element_text(size = 10)     # Adjust the legend text size
  ) +
  labs(
    title = "Grade of Esophageal Varix Before and After LAR-Octreotide",
    x = "Treatment Period",
    y = "Grade of Varices",
    fill = "Time Period",
    color = "Age Group"
  ) +
  stat_compare_means(
    aes(group = TimePeriod),                # Compare "Before" and "After" groups
    method = "wilcox.test",                 # Wilcoxon Signed-Rank Test
    paired = TRUE,                          # Paired test for before and after treatment
    label = "p.format",                     # Display exact p-value (e.g., p = 0.032)
    label.x = 1.5,                           # Position of the label
    label.y = 4.3
  )

# Display the plot
print(combined_boxplot_with_stats)

#### Boxplot with disease states ####
combined_boxplot_with_disease <- ggplot(processed_data_long, aes(x = TimePeriod, y = Grade_of_Varices, fill = TimePeriod)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_line(aes(group = Rand_id, color = AgeGroup), alpha = 0.5, size = 1) +
  geom_jitter(aes(color = TimePeriod), width = 0, size = 2, shape = 21, alpha = 0.7) +
  scale_fill_manual(values = c("Before" = "gray80", "After" = "gray50")) +
  scale_color_manual(values = c("Before" = "black", "After" = "darkgray")) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black"),
    axis.text = element_text(color = "black"),
    legend.position = "bottom"
  ) +
  labs(
    title = "Boxplot of Grade of Varices by Disease State with Individual Points Before and After Treatment",
    x = "Time Period",
    y = "Grade of Varices",
    fill = "Time Period"
  ) +
  facet_wrap(~ Disease.State) +  # Facet by Disease State
  stat_compare_means(
    aes(group = TimePeriod),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format",
    label.x = 1.5,
    label.y = 4.5
  )

print(combined_boxplot_with_disease)
```

# Statistical Analysis

```{r}

#### linear mixed models for the effects of age groups on grade of varices ####
# Rank-transform the Grade of Varices
processed_data_long$Rank_Grade_of_Varices <- rank(processed_data_long$Grade_of_Varices)

# Fit the linear mixed model with random effects for AgeGroup and Rand_id
lmm_model_age <- lmer(
  Grade_of_Varices ~ TimePeriod + (1 | AgeGroup) + (1 | Rand_id),
  data = processed_data_long
)
# Table with intercept and title
# Table with intercept and title
tbl_age <- tbl_regression(lmm_model_age, intercept = TRUE) %>%
  modify_caption("**Linear Mixed Model: Effects of Age Group on Varice Grade**")
print(tbl_age)

# Summarize the model (optional for your review)
print(summary(lmm_model_age))

# Fit the linear mixed model using Rank_Grade_of_Varices as the outcome
lmm_model_disease <- lmer(Grade_of_Varices ~ TimePeriod * Disease.State + (1 | Rand_id), data = processed_data_long)
tbl_disease <- tbl_regression(lmm_model_disease, intercept = TRUE) %>%
  modify_caption("**Linear Mixed Model: Effects of Disease State on Varice Grade**")
print(tbl_disease)

# Step 2: Display the summary of the model
summary(lmm_model_disease)



#### run Hb, MCV, PLTE, Alb, Na with linear mixed-effects model ####

# Step 1: Merge the two datasets
merged_data <- merge(datatable, labsdata, by.x = "rand", by.y = "rand_id", all.x = TRUE, all.y = TRUE)
merged_data$Date.of.Transplant <- as.Date(merged_data$Date.of.Transplant, format = "%m/%d/%Y")
merged_data$Date.of.Shunt <- as.Date(merged_data$Date.of.Shunt, format = "%m/%d/%Y")

# Step 2: Create a condition to exclude rows from labsdata
exclude_condition <- (
  # Only apply exclusion if both Date.of.Transplant and Date.of.Shunt are not missing
  (!is.na(merged_data$Date.of.Transplant) & merged_data$date.of.first.dose < merged_data$Date.of.Transplant) |
  (!is.na(merged_data$Date.of.Shunt) & merged_data$date.of.first.dose < merged_data$Date.of.Shunt)
) &
(
  # Only exclude if event_end_dt is after the corresponding Date.of.Transplant or Date.of.Shunt
  (!is.na(merged_data$Date.of.Transplant) & merged_data$event_end_dt > merged_data$Date.of.Transplant) |
  (!is.na(merged_data$Date.of.Shunt) & merged_data$event_end_dt > merged_data$Date.of.Shunt)
)


# Step 3: Filter out the specific rows from merged_data that meet the condition
merged_data_filtered <- merged_data[!exclude_condition, ]

# Step 4: Remove participants with only one data point
participant_counts <- table(merged_data_filtered$rand)
participants_to_remove <- names(participant_counts[participant_counts == 1])

# Filter out those participants with only one data point
merged_data_filtered <- merged_data_filtered[!merged_data_filtered$rand %in% participants_to_remove, ]

# Step 5: Check the filtered data
nrow(merged_data_filtered)  # Check how many rows are left after filtering
table(merged_data_filtered$rand)  # Check how many unique rand values remain

# Step 6: Run the model for each event_desc
event_desc_values <- c("HGB", "MCV", "PLTE", "Albumin", "Sodium") 
model_results <- list()

for (event in event_desc_values) {
  merged_data_filtered$result_val <- as.numeric(as.character(merged_data_filtered$result_val))
  merged_data_filtered$octreotide_status <- as.factor(merged_data_filtered$octreotide_status)
  merged_data_filtered$octreotide_status <- relevel(merged_data_filtered$octreotide_status, ref = "before")

  # Fit the model
  df <- merged_data_filtered %>% filter(event_desc == event)
  model <- lmer(result_val ~ octreotide_status + (1 | rand), data = df)
  conf <- confint(model)
  model_results[[event]] <- summary(model)

  # Save model summary using tbl_regression
  tbl <- tbl_regression(model, intercept = TRUE) %>%
  modify_caption(glue::glue("**Linear Mixed Model: {event}**"))
  model_results[[event]] <- tbl
  
  # Print the model summary for each event in a readable format
  print(paste("Results for", event, ":"))
  print(tbl)
  print(summary(model))
}
```

# Conclusion

A Wilcoxon signed-rank test was first used for assessing overall differences in varices grade before and after treatment, yielding a significant reduction (Figure 2, p = 0.014).

We used a linear mixed model with random intercepts for age group to assess the relationship between varices grade and octreotide status, finding a significant reduction in varices grade with treatment (p = 0.00639) while age-related variance was negligible (p \> 0.05). Similarly, although autoimmune hepatitis (p = 0.01620) and biliary atresia (p = 0.03888) were associated with higher varices grades, disease state did not significantly alter the effect of octreotide treatment (all p \> 0.05), suggesting minimal variability in treatment response due to age or disease state.

The linear mixed models showed significant changes in laboratory values during octreotide therapy. For hemoglobin (before treatment = 9.41), levels increased by 0.22 during therapy (p\<0.001) and by 0.57 after therapy (p\<0.001). MCV (before treatment = 85.73) decreased 0.83 during therapy (p\<0.001) and by 1.67 after therapy (p\<0.001). Platelet count (before treatment = 152.67) significantly decreased 45.47 during therapy (p\<0.001) and by 16.10 after therapy (p\<0.001). Albumin (before treatment = 3.27) increased 0.14 during therapy (p\<0.001) and by 0.28 after therapy (p\<0.001). Sodium (before treatment = 137.68) decreased 0.88 during therapy (p\<0.001) and by 0.46 after therapy (p\<0.001). These findings indicate significant, consistent effects of octreotide therapy on these clinical markers.  
